---
title: "The bad news about non-structural lack of positivity for continuous exposures"
format: gfm
bibliography: citations.bib
---

The *assignment mechanism* is key to any causal analysis. There are three assumptions: (1) the assignment is *individualistic*, that is an individual's assignment probability cannot depend on the values of covariates or potential outcomes of other individuals, (2) the assignment is *unconfounded*, the assignment mechanism cannot depend on the potential outcomes, and *probabilistic*, there is a non-zero probability for each treatment value for every individual. [@imbens2015causal] This paper focuses on the final assumption, sometimes referred to as *positivity*. Formally, we denote the continuous exposure as $T$, which takes on values in a set $\mathscr{T}$. For individual $i$ and each value of the exposure, $t$, there is a potential outcome $Y_i(t)$. [@imbens2000role] Each individual also has a vector of pre-exposure covariates, $X_i$. In the most general case, the assignment mechanism is defined as $P(T=t|\mathbf{X}, \mathbf{Y}(t))$ (with the unconfounded assumption this reduced to $P(T=t|\mathbf{X})$. An assignment mechanism is *probabilisitic* if the probability of assignment to each exposure level is strictly between 0 and 1:

$$0 < P_i(X, Y(t)) < 1, \textrm{ for each possible } X, Y(t)$$ 

for all $i=1, \dots, N$
In other words, the probabilistic (positivity) assumption states that for all individuals within strata of $X$ there exists a non-zero probability of receiving every exposure level. 

## GPS

A generalized propensity score is defined as:

$$e(t, x) = P(T = t | X = x)$$
Often, for continuous exposures this is estimated by first fitting a linear regression model prediction the exposure from a set of pre-exposure covariates. We then use the fitted values and the model variance in a Gaussian probability density function. [@austin2019assessing] 

$$e(t, x) = f_{T|X}(t|x) = \frac{1}{\sqrt{2\pi\hat\sigma^2}}\exp\left\{-\frac{(t-X\hat\beta)^2}{2\pi\hat\sigma^2}\right\}$$
The following stabilized weight is then used, where the numerator is the marginal density of the exposure.

$$w = \frac{f_T(t)}{f_{T|X}(t|x)} = \frac{\hat{\sigma}_{t|x}}{\hat\sigma_t}\exp\left\{\frac{(t-X\hat\beta)^2}{2\hat\sigma_{t|x}^2}-\frac{(t-\mu_t)^2}{2\hat\sigma_t^2}\right\}$$
## Positivity

Violations (or near violations) of the probabilistic assumption can increase both the bias and variance of causal effect estimates. [@petersen2012diagnosing]. There are two ways this violation can arise: structural non-positivity and random non-positivity. The former suggests that there is a structural mechanism that makes certain levels of the exposure impossible for a subset of individuals. The latter are random violations that can occur in finite samples due to chance. This paper will focus on random violations of positivity, demonstrating that "finite" does not mean small, but rather in the case of continuous exposures can be any sample size that is not infinite.

## Simulations

We examine 3 scenarios:

1. A single binary confounder with a moderate effect on the exposure and outcome
2. A single binary confounder with a large effect on the exposure and moderate effect on the outcome
3. A single binary confounder with a large effect on the exposure and large effect on the outcome.

We generate a binary confounder, $X$, from a Bernoulli distribution with probability 0.5. The continuous exposure, $T$ is generated as follows:

$$
T = aX + \varepsilon_{t|x}
$$ {#eq-t}
Where $\varepsilon_{t|x}\sim N(0,1)$ and $Y$ such that the "true" effect of $T$ is 0:

$$
Y = bX + \varepsilon_{y|x}
$$ {#eq-y}

where $\varepsilon_{y|x}\sim N(0,1)$. We examine $a = 1, 10$ and $b=1,10$ for moderate and large effects, respectively. We vary the sample size from 100 to 1,000,000, examining 100 replicates of each.

When $a = 1$, the positivity assumption is not violated, for example @fig-a1 shows a mirrored histogram for a simulation as specified above with a sample size of 10,000. When $a = 10$, however, we do see a near positivity violation, for all sample sizes. Here, because $T$ is continuous, the individuals at the population level *do* have a non-zero probability of receiving all other values of the exposure, however in practice it happens very infrequently. See @fig-a2, a mirrored histogram for a simulation where $a=10$ and a sample size of 10,000.

```{r}
#| label: "fig-a1"
#| fig-cap: "Mirrored Histogram showing overlap. a = 1, b = 1, n = 10,000"
#| echo: false
library(targets)
tar_read(fig_a1)
```

```{r}
#| label: fig-a2
#| fig-cap: "Mirrored histogram with positivity near-violation"
#| echo: false
tar_read(fig_a2)
```

```{r}
#| label: fig-weight
#| fig-cap: "oh no. Weighted pseudopopulation using GPS (ATE) weights"
#| echo: false
tar_read(fig_weight)
```

The estimates for the effect of the treatment are biased. The distribution for the effect is non-normal (@fig-skew).

```{r}
#| label: fig-skew
#| fig-cap: "Skewed distribution of the estimated coefficient for the treatment. For reference a normal density is overlaid in orange."
#| echo: false
tar_read(fig_skew)
```

@fig-sims demonstrates that in the case of a continuous exposure, when there is near violation of the positivity assumption, as seen in panels B and C, the causal estimate using the stabilized generalized propensity score is both biased and has large variance, which does not appreciably decrease despite the large sample size.


```{r}
#| label: fig-sims
#| fig-height: 10
#| message: false
#| warning: false
#| fig-cap: "Boo. Look at all that bias and variance"
#| echo: false
tar_read(fig_sims)
```


The magnitude of the bias and variability depend on the magnitude of the effect of $X$ and $T$ (in @eq-t as $a$), and, for a binary confounder the prevalence of the confounder ($p$). Below is a heatmap exploring the relationship between these two quantities when $n = 10,000$

```{r}
#| label: fig-bias
#| message: false
#| warning: false
#| fig-cap: "Impact of the prevalence of X, magnitude of the effect between X and T, and magnitude of the effect between X and Y on the bias of the observed exposure effect."
#| echo: false
tar_read(fig_bias)
```


```{r}
#| label: fig-variance
#| fig-cap: "Impact of the prevalence of X, magnitude of the effect between X and T, and magnitude of the effect between X and Y on the variability in the observed exposure effect."
#| echo: false
tar_read(fig_variance)
```

